Ownership types in theory and practice (in Rust)

### 2023-01-18 paper/paper.tex
Expanded owners-as-modifiers section

\usepackage{booktabs} % For formal tables
\usepackage{subcaption}
\usepackage{tikz}
\usetikzlibrary{fit}
\usepackage{pgfplots}
\usepackage{pgfplotstable}
\usepackage{hyphenat}
\usepackage{todonotes}
\usepackage[babel]{csquotes}
\usepackage{listings}
\lstset{frame=tb,
  % language=Rust,
  aboveskip=3mm,
  belowskip=3mm,
  showstringspaces=false,
  columns=flexible,
  basicstyle={\small\ttfamily},
  numbers=left,
  numbersep=5pt,
  xleftmargin=\parindent,
  numberstyle=\color{gray},
  % keywordstyle=\color{blue},
  % commentstyle=\color{dkgreen},
  % stringstyle=\color{mauve},
  breaklines=true,
  breakatwhitespace=true,
  tabsize=2
}
\pagenumbering{gobble}

@@

We initialize a vector \emph{v} that contains two integers stored in a buffer in memory.
Next, we create a pointer \emph{vptr} that points into this buffer, specifically to the place where the second element (with current value 11) is stored.
Now, both \emph{v} and \emph{vptr} point to (overlapping parts of) the same buffer.
% TODO: change wording from source
We say that the two pointers are aliasing.
Then, we push a new element to the end of \emph{v}.
If the vector's capacity were large enough, the new element would be appended to the buffer.
However, we will assume there is no more space for an additional element, so a new buffer is allocated, and all the existing elements are moved over.
This case is interesting because \emph{vptr} still points to the old buffer!
In other words, adding a new element to \emph{v} has turned \emph{vptr} into a dangling pointer.
Therefore, trying to access the value the dangling pointer \emph{vptr} points to will cause a use-after-free bug.
% More generally, we can observe that an action through a pointer (\emph{v}) will also affect all of its aliases (\emph{vptr}).
The main problem here is that such memory issues cannot be detected at compile time.

@@

@@

The core concept of ownership types is that objects can be \emph{owners} of other objects.
Given the two objects \emph{a} and \emph{b}, \emph{a} is \emph{inside} \emph{b} if \emph{a} is the same object as \emph{b} or \emph{a} is \emph{owned} by \emph{b}, transitively.
% Given the two objects \emph{a} and \emph{b}, \emph{a} is \emph{outside} \emph{b} if \emph{b} is the same object as \emph{a} or \emph{b} is \emph{owned} by \emph{a}, transitively.
The \emph{outside} relation is the converse of the \emph{inside} relation.

@@

Yang Zhao and John Boyland \cite{permission-ownership-types} specify that, with owners-as-dominators, each regular object is owned by exactly one object and can be the owner of zero or more objects.
Objects can also be owned by a special global object called \emph{world}, which cannot be owned anymore.
This \emph{world} object is the root of the acyclic ownership hierarchy tree that represents the ownership relation.
In this tree structure, each object is \emph{inside} its owner.
This model is named \emph{owners-as-dominators} because any object acting as an owner is considered to be a \emph{dominator} for all of its owned objects.

@@

    % TODO: make green darker with:
    % \definecolor{mynicegreen}{RGB}{102,252,102}

    \draw[->,thick,dashed,green] (Car) edge[bend left] node[right] {owner} (Engine);
    \draw[->,thick,dashed,green] (Engine) edge[bend left] node[left] {inside} (Car);
    \draw[->,thick,dashed,green] (Car) edge node[above] {sibling} (Driver);
    \draw[->,thick,dashed,red] (World) edge[bend right=90] node[left] {invalid} (Engine);
    \draw[->,thick,dashed,red] (Driver) edge[bend left] node[right] {invalid} (Engine);
  \end{tikzpicture}

@@

The object \emph{engine} is \emph{inside} the object \emph{car} because \emph{engine} is owned by \emph{car}.
Conversely, \emph{car} is outside \emph{engine}.
Therefore, \emph{engine} can validly reference \emph{car} according to the third rule.

Even the original Flexible Alias Protection model, which these ownership types are based on, imposed the restriction "that the only objects that can access [an object] are the object that owns it and other objects inside that [object]" \cite{flexible-alias-protection}.
The implication is that this model only protects objects from external access.
Internally, an object can still access its owner (and its owner's owner etc.).

@@

@@

Most importantly, these three rules imply that any external reference to an object is only possible through its owner.
This is also examplified in Figure \ref{fig:owners-as-dominators} where the object \emph{world} is trying to access the object \emph{engine}.
This is not allowed in the owners-as-dominators model, as \emph{world} is neither a direct owner, a sibling or inside of \emph{engine}.
Instead, \emph{world} could only reference \emph{engine} through the \emph{engine}'s owner \emph{car}.

Its strictness can be seen as both an advantage and disadvantage of the owners-as-dominators model.
% TODO: reference ownership survey better, as this is nearly copied
The advantage is that the model provides a simple, clear and strong guarantee that allows for reasoning about various properties of the code.
% TODO: provide example of advantages
However, it also makes programming more difficult by not allowing common idioms that involve aliasing.
% TODO: example of such idiom that isn't representable

% TODO: Add rust section on taking ownership of fields inside type
% Struct Car {
%   engine: Engine,
% }

% let car = Car::new(Engine::new());
% let driver = Driver::new();

@@

Another kind is the \emph{owners-as-modifiers} model \cite{ownership-types-survey}, which is a weaker form of owners-as-dominators that additionally allows read-only references.
A read-only reference can only be used to read files and call pure methods, which may not modify any existing objects.
In essence, aliasing is unrestricted, but only the owner of an object can modify it.
Owners-as-modifiers extend the rules specified for owners-as-dominators.

% For \emph{a} to validly reference \emph{b} through reference \emph{r}, owners-as-modifiers follows the same rules as owners-as-dominators.
% The only addition is that \emph{r} can also be a valid reference if it is a read-only reference and only pure methods are being called on it.
For \emph{a} to validly reference \emph{b} through reference \emph{r}, owners-as-modifiers follows the following rules \cite{ownership-types-survey}.
\begin{enumerate}
  \item \emph{a} is the owner of \emph{b},
  \item \emph{a} and \emph{b} are siblings, or
  \item \emph{b} is outside of \emph{a}, or
  \item \emph{r} is a read-only reference and only pure methods can be called on it.
\end{enumerate}

According to Werner Dietl and Peter Müller \cite{lightweight-ownership}, a fundamental difference between owners-as-dominator and owners-as-modifiers is that the former restricts \emph{where} references are allowed to point to, while the latter allows references to point to objects in arbitrary contexts, but restricts \emph{how} references can be used.

The owners-as-modifiers principle expands the programming possibilities of the owners-as-modifiers principle and fulfills the requirements needed for the verification of the functional correctness of object-oriented programs.

@@

@@

Heap-based values will always be generated by a function at runtime.
Therefore, one can generalize the above scenario to ownership being transferred from the variable \emph{a} to the variable \emph{b} when \emph{b} is assigned to the return value \emph{a} of \emph{foo}.
\begin{lstlisting}
fn main() {
  let b = foo();
}
fn foo() {
  let a = HeapBasedType::default();
  a
}
\end{lstlisting}

@@

In contrast to heap-based values, the size of stack-based values is known at compile time and is usually small.
Therefore, stack-based values in Rust are \emph{copied} instead of \emph{moved}.
Internally, this works because primitive types are annotated with and implement the \emph{Copy} trait.

@@

\paragraph{Limits of the ownership approach}

>>>>

\usepackage{booktabs} % For formal tables
\usepackage{subcaption}
\usepackage{tikz}
\usetikzlibrary{fit}
\usepackage{pgfplots}
\usepackage{pgfplotstable}
\usepackage{hyphenat}
\usepackage{todonotes}
\usepackage[babel]{csquotes}
\usepackage{listings}
\lstset{frame=tb,
  % language=Rust,
  aboveskip=3mm,
  belowskip=3mm,
  showstringspaces=false,
  columns=flexible,
  basicstyle={\small\ttfamily},
  numbers=left,
  numbersep=5pt,
  xleftmargin=\parindent,
  numberstyle=\color{gray},
  % keywordstyle=\color{blue},
  % commentstyle=\color{dkgreen},
  % stringstyle=\color{mauve},
  breaklines=true,
  breakatwhitespace=true,
  tabsize=2
}
\pagenumbering{gobble}
\definecolor{cgreen}{RGB}{60,201,65}

@@

We initialize a vector \emph{v} that contains two integers stored in a buffer in memory.
Next, we create a pointer \emph{vptr} that points into this buffer, specifically to the place where the second element (with current value 11) is stored.
Now, both \emph{v} and \emph{vptr} point to (overlapping parts of) the same buffer.
% TODO: change wording from source
We say that the two pointers are aliasing.
Then, we push a new element to the end of \emph{v}.
If the vector's capacity were large enough, the new element would be appended to the buffer.
However, we will assume there is no more space for an additional element, so a new buffer is allocated, and all the existing elements are moved over.
This case is interesting because \emph{vptr} still points to the old buffer!
In other words, adding a new element to \emph{v} has turned \emph{vptr} into a dangling pointer.
Therefore, trying to access the value which the dangling pointer \emph{vptr} points to will cause a use-after-free bug.
More generally, we can observe that an action through a pointer (\emph{v}) will also affect all of its aliases (\emph{vptr}), even though these aliases might not expect a change.
The main problem here is that such memory issues cannot be detected at compile time.

@@

In object-oriented programs, an object can reference any other object and read and modify its fields through direct field accesses or method calls.
Such programs with arbitrary object structures are difficult to understand, to maintain, and to reason about. \cite{lightweight-ownership}

Ownership types can limit which objects can be referenced and can specify whether the referenced objects may be mutated or just read from.

@@

The core concept of ownership types is that objects can be \emph{owners} of other objects.

Given the two objects \emph{a} and \emph{b}, \emph{a} is \emph{inside} \emph{b} if \emph{a} is the same object as \emph{b} or \emph{a} is \emph{owned} by \emph{b}, transitively.
The \emph{outside} relation is the converse of the \emph{inside} relation.

@@

Yang Zhao and John Boyland \cite{permission-ownership-types} specify that, with owners-as-dominators, each regular object is owned by exactly one object and can be the owner of zero or more objects.
Objects can also be owned by a special global object called \emph{world}, which cannot be owned anymore.
This \emph{world} object is the root of the acyclic ownership hierarchy tree that represents the ownership relation.
In this tree structure, each object is inside its owner.
This model is named \emph{owners-as-dominators} because any object acting as an owner is considered to be a \emph{dominator} for all of its owned objects.

@@

    \draw[->,thick,dashed,cgreen] (Car) edge[bend left] node[right] {owner} (Engine);
    \draw[->,thick,dashed,cgreen] (Engine) edge[bend left] node[left] {inside} (Car);
    \draw[->,thick,dashed,cgreen] (Car) edge node[above] {sibling} (Driver);
    \draw[->,thick,dashed,red] (World) edge[bend right=90] node[left] {invalid} (Engine);
    \draw[->,thick,dashed,red] (Driver) edge[bend left] node[right] {invalid} (Engine);
  \end{tikzpicture}

@@

@@

The object \emph{engine} is inside the object \emph{car} because \emph{engine} is owned by \emph{car}.
Conversely, \emph{car} is outside \emph{engine}.
Therefore, \emph{engine} can validly reference \emph{car} according to the third rule.

The original Flexible Alias Protection model, which these ownership types are based on, also imposed the restriction "that the only objects that can access [an object] are the object that owns it and other objects inside that [object]" \cite{flexible-alias-protection}.
The implication is that this model only protects objects from external access.
Internally, an object can still access its owner (and its owner's owner etc.).

% TODO: why iterators can be implemented because of 3. outside rule
% This rule may seem quite unintuitive, which is underlined by the fact that Dietl et al.'s version of owners-as-dominators leave this rule out.

% Dietl et al. claim that Clarke's owners-as-dominators property is weakened "by allowing instances of inner classes to access the representation of the instance of the outer class they are associated with.
% Dietl et al. claim that, "by allowing instances of inner classes to access the representation of the instance of the outer class they are associated with, [...] they can handle iterators, but not more general forms of sharing."
% parametric ownership type systems describe ownership properties accurately and
% guarantee a strong type invariant, ownership parametricity increases the complexity
% of the type system and the annotation overhead.

@@

Primarily, these three rules imply that any external reference to an object is only possible through its owner.
This is examplified in Figure \ref{fig:owners-as-dominators} where the object \emph{world} is trying to access the object \emph{engine}, which is not allowed because \emph{world} is neither a direct owner of, a sibling of or inside of \emph{engine}.
Instead, \emph{world} could only reference \emph{engine} through the \emph{engine}'s owner \emph{car}.

According to Dietl et al. \cite{lightweight-ownership}, the owners-as-dominators model restricts \emph{where} references are allowed to point to.
% TODO: find reference/quote
Any object that can validly be referenced may be mutated through that reference.

Its strictness can be seen as both an advantage and a disadvantage of the owners-as-dominators model.
The advantage is that the model provides a simple, clear and strong guarantee that allows for reasoning about various properties of the code.
However, it also makes programming more difficult by not allowing common idioms that involve aliasing. \cite{ownership-types-survey}

% TODO: mention that all objects that can be referenced can be mutated

% TODO: Add rust section on taking ownership of fields inside type
% Struct Car {
%   pub engine: Engine,
% }

% let car = Car::new(Engine::new());
% let driver = Driver::new();
% let engine = car.engine;

@@

Another kind is the \emph{owners-as-modifiers} model \cite{ownership-types-survey}, which is a weaker form of owners-as-dominators that additionally allows read-only references.
A read-only reference can only be used to read files and call \emph{pure methods}.
Pure methods are defined as methods that do not modify the fields of any existing objects. \cite{ownership-types-survey}

For \emph{a} to validly reference \emph{b} through reference \emph{r}, at least one of the following rules \cite{ownership-types-survey} must apply.
\begin{enumerate}
  \item \emph{a} is the owner of \emph{b},
  \item \emph{a} and \emph{b} are siblings, or
  \item \emph{b} is outside of \emph{a}, or
  \item \emph{r} is a read-only reference and only pure methods can be called on it.
\end{enumerate}
We can see that the owners-as-modifiers concept extends the rules specified for owners-as-dominators.
The only additional rule is that a reference can also be valid if it is a read-only reference and only pure methods are being called on it.

    \draw[->,thick,dashed,cgreen] (Car) edge[bend left] node[right] {owner} (Engine);
    \draw[->,thick,dashed,cgreen] (Engine) edge[bend left] node[left] {inside} (Car);
    \draw[->,thick,dashed,cgreen] (Car) edge node[above] {sibling} (Driver);
    \draw[->,thick,dashed,orange] (World) edge[bend right=90] node[left] {read only} (Engine);
    \draw[->,thick,dashed,orange] (Driver) edge[bend left] node[right] {read only} (Engine);
  \end{tikzpicture}

  \caption{Owners-as-modifiers \cite{ownership-types-survey}: Solid lines indicate "owns" and dotted lines indicate "references".}
  \label{fig:owners-as-modifiers}
\end{figure}

\paragraph{Implications}

Dietl et al. \cite{lightweight-ownership} suggest that the owners-as-modifiers approach allows references to point to objects in arbitrary contexts, but restricts \emph{how} references can be used.
In particular, any object can be referenced by any given object.
However, only an object that is the owner of, a sibling of or inside of another given object may receive a reference to that object that allows mutating the object.
Figure \ref{fig:owners-as-modifiers} demonstrates that all other objects, such as \emph{world} or \emph{driver}, which may not be allowed to obtain a mutable reference to an object for the reasons above, can still receive read access to said object.

% TODO: find better disadvantage
A disadvantage of the model may be the added complexity of read-only references.
However, its advantage is that it expands the programming possibilities of the owners-as-dominators principle and fulfills the requirements needed for the verification of the functional correctness of object-oriented programs.

@@

One common restriction to the early ownership systems is that the owner of an object must be set upon creation and then fixed for the lifetime of the object \cite{ownership-types-survey}.
Rust extends the original ownership concepts by allowing the ownership of an object to be transferred.

@@

Heap-based values will always be generated by a function at runtime.
In the above example, the return value of the function \verb|HeapBasedType::default()| is moved into \emph{a}.
Furthermore, one can generalize the above scenario to ownership being transferred from the variable \emph{a} to the variable \emph{b} when \emph{b} is assigned to the return value \emph{a} of \emph{foo}.
\begin{lstlisting}
fn main() {
  let b = foo();
}
fn foo() {
  let a = HeapBasedType::default();
  a
}
\end{lstlisting}

@@

In contrast to heap-based values, the size of stack-based values is known at compile time and is usually small.
Therefore, stack-based values in Rust are \emph{copied} instead of \emph{moved}.
% TODO: explain copy trait
Internally, this works because primitive types are annotated with the \emph{Copy} trait.

@@

% \paragraph{Limits of the ownership approach}
